---
title: "Lepto_iNat_Pops"
format: pdf

execute:
  warning: false
---

# Babystars and soil salinity

Anthocyanins are a group of pigments that contribute to petal color intensity across many species of flowers. These pigments also are important for providing tolerance to abiotic stressors, such as high UV radiation, extreme soil conditions, and low precipitation.

*Leptosiphon minimus* (common name: True babystars) is a native plant species found across the northwestern coast of North America in California, Oregon, Washington, and British Columbia. *L. minimus* is known for being polymorphic, exhibiting two petal colors: pink and white. Within these color morphs, elevated levels of anthocyanins produce a pink pigment.

In our project, we are interested in assessing if certain abiotic stressors, specifically harsh, more saline soils, can predict the frequency of pink flowers for *L. minimus* at the population level. We will be using iNaturalist observations taken across the northwestern coast of USA and Canada to identify the pink vs. white color morphs of *L. minimus* and group them by different populations. We will also use distance to shoreline as a proxy for soil salinity.

**Question:** Are *L. minimus* flowers more likely to be pink the closer they are located to the shore?

**Hypothesis:** Soils closer to the sea shore have higher salinity, and increasing anthocyanin levels can help plants better withstand exposure to more saline microenvironments.

**Prediction:** If anthocyanins help plants better withstand exposure to saline soils, then we predict that *L. minimus* flowers closer to the sea shore will be more likely to be pink.

------------------------------------------------------------------------

## Set up Working Directory and Install/Library Packages

Let's start by loading the required packages (and installing if necessary).

```{r}
#install.packages("dbscan")
#install.packages("sf")
#install.packages(("ggspatial"))
#install.packages("rnaturalearthdata")
#install.packages("scatterpie")
#install.packages("rnaturalearth")
#install.packages("sf")
#install.packages("tmap")

library(scatterpie)
library(dbscan)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggspatial)
library(rnaturalearth)
library(tmap) # To visualize shoreline shape files
library(brms) # For Bayesian Statistical Modeling
library(ggeffects) 

```

------------------------------------------------------------------------

## Import data set into R

We have downloaded 69 observations of *Leptosiphin minimus* (babystars) from iNaturalist as a CSV file (called `Lmini_iNat.csv`). This will be our raw data, and includes the following columns: the observation ID (`id`), the URL to the observation on iNaturalist.com (`url`), latitude/longitude for the observation location (`latitude` and `longitude`) and the scientific name.

```{r}
#Read in CSV
LM <- read.csv("data/LMini_iNat.csv")

#Check out names of columns
names(LM)
```

------------------------------------------------------------------------

## Convert from Degrees to Meters, set up Pops

In order to define populations, we will use a package called `dbscan` to group *L. minimus* observations into populations using a 1000-meter radius. Before we can do this, we must convert latitude/longitude degrees into meters.

```{r}
#Convert to sf object
LM_sf <- st_as_sf(LM, coords = c("longitude", "latitude"), crs = 4326)

#Project to meters
LM_utm <- st_transform(LM_sf, 32611)  # change UTM zone if needed
LM_coords <- st_coordinates(LM_utm)
```

------------------------------------------------------------------------

## Run `dbscan` and add define populations

Now that we've converted latitude to meters, we can run dbscan. We will use dbscan to define populations as 4 or more observations within a 1000 m radius. 

```{r}
#Run DBSCAN, 1000 m radius populations with 4 or more observations per pop
db <- dbscan(LM_coords, eps = 1000, minPts = 4)  

#Add population column
LM$population_id <- db$cluster

```

------------------------------------------------------------------------

## Check if it worked

Let us check if we added a column for population ID. 

```{r}
table(LM$population_id)
sum(LM$population_id == 0)

```

------------------------------------------------------------------------

## Save CSV

```{r}
write.csv(LM, "data/LMini_iNat_Pops.csv", row.names = FALSE)
```

------------------------------------------------------------------------

## Visualize data

### NOTE: All observations beloning to population '0' are singelton observations, that did not meet our population criteria (4 obs within 1000 m)

Let's count how many observations are within each of populations 0-6. 

```{r}
#List pop sizes
LM_pop_sizes <- LM %>%
  count(population_id, name = "n_obs") %>%
  arrange(desc(n_obs))

LM_pop_sizes

#Bar plot of pop sizes
ggplot(LM_pop_sizes, aes(x = factor(population_id), y = n_obs)) +
  geom_col() +
  labs(x = "Population ID",
       y = "Number of Observations",
       title = "L. minimus Observation Counts per Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5))


```

------------------------------------------------------------------------

## Visualize pops on a map

```{r}
# Get world basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# Bounding box around our data
bbox <- c(
  xmin = min(LM$longitude) - 0.5,
  xmax = max(LM$longitude) + 0.5,
  ymin = min(LM$latitude) - 0.5,
  ymax = max(LM$latitude) + 0.5
)

# Plot
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_point(data = LM,
             aes(x = longitude, y = latitude,
                 color = factor(population_id)),
             size = 2, alpha = 0.85) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           expand = FALSE) +
  labs(color = "Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5))

```

## Lets remove noise from 'pop 0' singelton observation

```{r}
# Remove noise points (population_id == 0)
LM_clean <- LM %>%
  filter(population_id != 0)

# Quick check
table(LM_clean$population_id)

# Bounding box around our data
bbox <- c(
  xmin = min(LM_clean$longitude) - 0.5,
  xmax = max(LM_clean$longitude) + 0.5,
  ymin = min(LM_clean$latitude) - 0.5,
  ymax = max(LM_clean$latitude) + 0.5
)

#Plot cleaned data
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_point(data = LM_clean,
             aes(x = longitude, y = latitude,
                 color = factor(population_id)),
             size = 2, alpha = 0.85) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           expand = FALSE) +
  labs(color = "Population") +
  theme_minimal() +
   theme(axis.text.x = element_text(size = 5))

```

------------------------------------------------------------------------

## Finish and Save

```{r}
# Save to new CSV
write.csv(LM_clean, "data/LMini_iNat_Pops_filtered.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# Curating Data

------------------------------------------------------------------------

### Now that we have our data set, we need to mark iNat observations based on flower color

Lets make a copy of **LMini_iNat_Pops_filtered.csv** called **LMini_iNat_Pops_filtered_curated.csv**, and add some new columns: **polymorphic** and **color1** and **color2**.

```{r}
LMini_iNat_Pops_filtered = read.csv("data/LMini_iNat_Pops_filtered.csv")
LMini_iNat_Pops_filtered_curated <- LMini_iNat_Pops_filtered %>%
  mutate(
    polymorphic = NA,
    color1 = NA_character_,
    color2 = NA_character_
  )

write.csv(LMini_iNat_Pops_filtered_curated,
          "data/LMini_iNat_Pops_filtered_curated.csv",
          row.names = FALSE)
```

Next, we will go through each iNat observation in our LMini_iNat_Pops_filtered_curated.csv by clicking the hyperlink in each respective row. Then we will note the flower color in the observation's photo, choosing **pink** or **white**, and writing that down in the cell of the color column for each respective row. However, if the observation has no flowers or all flowers are still closed, we will instead write **closed** in that cell instead. Additionally, we will mark whether the observation show more than one flower color in the polymorphic column. This will be a binary variable, with **polymorphic observations receiving a 1** and **monomorphic observations receiving a 0.** If the observation is polymorphic, it's second color will be put into the color2 column.

Once the **polymorphic** and **color** columns are filled out we can move on to the next step; lets save the new file as **LMini_iNat_Pops_filtered_curated_color.csv** Make sure to save it as a CSV after editing.\

------------------------------------------------------------------------

## Filtering out closed flowers

```{r}
#Read in curated CSV 
LMC <- read.csv("data/LMini_iNat_Pops_filtered_curated_color.csv")

LMCF <- LMC %>%
  filter(!(color1 == "closed" & color2 == "closed"))
```

```{r}
#Combine remaining color columns into one column and remove empty rows
LMCF_clean <- LMCF %>%
  pivot_longer(
    cols = starts_with("color"),
    values_to = "color",
    names_to = "slot"
  ) %>%
  filter(
    !is.na(color),
    color != "",
    color != "closed"
  )
```

------------------------------------------------------------------------

## Group colors by population

```{r}
#Grouping colors by population for pie charts
pop_pies <- LMCF_clean %>%
  group_by(population_id) %>%
  summarise(
    latitude  = mean(latitude),
    longitude = mean(longitude),
    white = sum(color == "white"),
    pink  = sum(color == "pink"),
    total_n = n(),
    .groups = "drop"
  )
```

------------------------------------------------------------------------

## Mapping populations as pie charts

```{r}
# World basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# Bounding box from population-level data
bbox <- c(
  xmin = min(pop_pies$longitude) - 0.5,
  xmax = max(pop_pies$longitude) + 0.5,
  ymin = min(pop_pies$latitude) - 0.5,
  ymax = max(pop_pies$latitude) + 0.5
)

# Plot
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_scatterpie(
    data = pop_pies,
    aes(x = longitude, y = latitude),
    cols = c("white", "pink"),
    pie_scale = 1.5,
    color = "black"
  ) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = c(white = "white", pink = "#d81b60"),
    name = "Flower color"
  ) +
  theme_minimal()
```

## Mapping distance from shore

In order to calculate the distance of each observation from the water (i.e., to the closest shore), we will compute the minimum distance from each observation (point) to the shoreline using vectors from the NOAA National Shoreline CUSP (West region) dataset. We downloaded the West region shape file from the NOAA portal (<https://nsde.ngs.noaa.gov/>), and imported this into R. **Note: This only works for observations in the US, so we have to manually calculate the distance from shore for the Canadian observations (population 1) using Avenza.**

This will output a data frame with

```{r}

# Import shape file of Western US shoreline
shoreline <- st_read("data/West/West.shp")

# To view the imported shape file
tmap_mode("view")  # interactive mode
tm_shape(shoreline) + tm_lines()

# Import coordinate data for observations
points_sf <- st_as_sf(LMCF_clean,
                      coords = c("longitude", "latitude"),
                      crs = 4326)

# Project to a suitable projection
shoreline_proj <- st_transform(shoreline, 32610) # UTM 10N
points_proj    <- st_transform(points_sf, 32610)

# Compute minimum distances to shore - add new column called dist_m
LMCF_clean$dist_m <- apply(st_distance(points_proj, shoreline_proj), 1, min)

# Manually overwrite data for observations in Canada (Population 1)
LMCF_clean$dist_m[6] <- 200 # Observation 121793064
LMCF_clean$dist_m[7] <- 210 # Observation 121793096
LMCF_clean$dist_m[8] <- 190 # Observation 121793133

# Change pink to 1 and white to 0, add a new column called color_binom
LMCF_clean <- LMCF_clean %>%
  mutate(color_binom = case_when(
    color == "pink"  ~ 1,
    color == "white" ~ 0,
    TRUE ~ NA_real_
  ))

# Write to csv
write.csv(LMCF_clean,"data/LMCF_clean_dist.csv", row.names = FALSE)

# Check headers for data frame
str(LMCF_clean)
```

------------------------------------------------------------------------

## Make a simple ggplot for our potential model 

```{r}
# Change color to a factor - will default 
#LCMF_clean$color <- as.factor(LCMF_clean$color$)

# Plot 
ggplot(data=LMCF_clean, aes(x=dist_m, y=color)) +
  geom_point() +
  theme_bw()
```


## Run prior predictive simulations
```{r}
# Define priors
priors <- c(
  set_prior("normal(0, 1)", class = "Intercept"),        # prior for alpha (intercept)
  set_prior("normal(0, .5)", class = "b", coef = "dist_m"), # prior for beta (slope)
  set_prior("exponential(1)", class = "sd")                 # prior for random effect SD
)

# Run prior simulations
n <- 1000

priorsims <- tibble(
    group = seq_len(n),
    alpha = rnorm(n, 0, 1),   # prior for intercept (log-odds scale)
    beta  = rnorm(n, 0, .5)      # prior for slope (effect of distance)
  ) %>%
  expand(
    nesting(group, alpha, beta),
    dist_scaled = seq(from = -2, to = 2, length.out = 100)  # range of (scaled) distance
  ) %>%
  mutate(
    log_odds = alpha + beta * dist_scaled,   # linear predictor (log-odds)
    prob      = plogis(log_odds)             # convert to probability via inverse logit
  )

# Plot prior predictive lines
ggplot(priorsims,
       aes(x = dist_scaled,
           y = prob,
           group = group)) +
  geom_line(alpha = 0.05) +
  labs(
    title = "Prior Predictive Lines (Bernoulli logit)",
    x     = "Distance to Shore (centered & scaled)",
    y     = "P(pink flower)"
  )
```

## Run the model with our priors
```{r}
# Run the model
m.dist.color.pop <-
  brm(data = LMCF_clean, # Give the model the data
      # Use a binomial distribution
      family = bernoulli(link = "logit"),
      # Specify the model 
      color_binom ~ 1 + dist_m + (1|population_id),
      # Set priors 
      prior = priors, 
      sample_prior = TRUE,
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      file = "output/m.dist.color.pop")
```

## Assess the model

Let's run summary(), print(), and plot() on our model output to assess how well it ran! 

The model seems to have converged correctly! Our Rhat is equal to 1. The MCMC chains are overlapping and flat, and the posterior distributions are smooth and have one clean peak. 

For every 1 meter increase in distance to shore, the log-odds of being pink increases by 0.003. However, we cannot confidently state that the effect size is different from 0 because our 95% lower and upper compatibility intervals intersect 0 (-0.001 - 0.010). 
```{r}
summary(m.dist.color.pop)
print(m.dist.color.pop, digits = 3)
plot(m.dist.color.pop)
```
## How often is our posterior slope larger than zero? 

In this section, we will use the as_draws_df() to extract the posterior distribution samples from our fitted brms model into a data frame. Then, we will calculate a proportion of how many our sampled values of our slope are positive. 
```{r}
#  Extract all posteriors from 4 MCMC chains into a table
draws <- as_draws_df(m.dist.color.pop) 

# What proportion of posteriors have estimate for slope > 0?
sum(draws$b_dist_m>0)/length(draws$b_dist_m) 
```

We have a value of .858, which means that 85.8% of the posterior distribution for the slope is above zero. This means that there is an 85.8% probability that distance to shore has a positive effect on pink flower color. In other words, flowers are more likely to be pink as distance from shore increases. 


```{r}
# Plot predicted response
preds <- predict_response(m.dist.color.pop,
                          bias_correction = TRUE)
plot(preds)
```
