---
title: "Lepto_iNat_Pops"
format: pdf

execute:
  warning: false
---

------------------------------------------------------------------------

## Set up Working Directory and Install/Library Packages

```{r}
#install.packages("dbscan")
#install.packages("sf")
#install.packages(("ggspatial"))
#install.packages("rnaturalearthdata")
#install.packages("scatterpie")
#install.packages("rnaturalearth")
#install.packages("sf")

library(scatterpie)
library(dbscan)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggspatial)
library(rnaturalearth)
```

\

------------------------------------------------------------------------

# *Leptosiphon minimus*

------------------------------------------------------------------------

## Read in CSV from iNat (*Leptosiphon minimus*)

```{r}
#Read in CSV
LM <- read.csv("data/LMini_iNat.csv")

#Check out names of columns
names(LM)
```

------------------------------------------------------------------------

## Convert from Degrees to Meters, set up Pops

```{r}
#Convert to sf object
LM_sf <- st_as_sf(LM, coords = c("longitude", "latitude"), crs = 4326)

#Project to meters
LM_utm <- st_transform(LM_sf, 32611)  # change UTM zone if needed
LM_coords <- st_coordinates(LM_utm)
```

------------------------------------------------------------------------

## Run dbscan and add Pops column to CSV

```{r}
#Run DBSCAN, 1000 m radius populations with 4 or more observations per pop
db <- dbscan(LM_coords, eps = 1000, minPts = 4)  

#Add population column
LM$population_id <- db$cluster

```

------------------------------------------------------------------------

## Check if it worked

```{r}
table(LM$population_id)
sum(LM$population_id == 0)

```

------------------------------------------------------------------------

## Save CSV

```{r}
write.csv(LM, "data/LMini_iNat_Pops.csv", row.names = FALSE)
```

------------------------------------------------------------------------

## Visualize data

### NOTE: All observations beloning to population '0' are singelton observations, that did not meet our population criteria (4 obs within 1000 m)

```{r}
#List pop sizes
LM_pop_sizes <- LM %>%
  count(population_id, name = "n_obs") %>%
  arrange(desc(n_obs))

LM_pop_sizes

#Bar plot of pop sizes
ggplot(LM_pop_sizes, aes(x = factor(population_id), y = n_obs)) +
  geom_col() +
  labs(x = "Population ID",
       y = "Number of Observations",
       title = "L. minimus Observation Counts per Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5))


```

------------------------------------------------------------------------

## Visualize pops on a map

```{r}
# Get world basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# Bounding box around our data
bbox <- c(
  xmin = min(LM$longitude) - 0.5,
  xmax = max(LM$longitude) + 0.5,
  ymin = min(LM$latitude) - 0.5,
  ymax = max(LM$latitude) + 0.5
)

# Plot
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_point(data = LM,
             aes(x = longitude, y = latitude,
                 color = factor(population_id)),
             size = 2, alpha = 0.85) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           expand = FALSE) +
  labs(color = "Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 5))

```

## Lets remove noise from 'pop 0' singelton observation

```{r}
# Remove noise points (population_id == 0)
LM_clean <- LM %>%
  filter(population_id != 0)

# Quick check
table(LM_clean$population_id)

# Bounding box around our data
bbox <- c(
  xmin = min(LM_clean$longitude) - 0.5,
  xmax = max(LM_clean$longitude) + 0.5,
  ymin = min(LM_clean$latitude) - 0.5,
  ymax = max(LM_clean$latitude) + 0.5
)

#Plot cleaned data
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_point(data = LM_clean,
             aes(x = longitude, y = latitude,
                 color = factor(population_id)),
             size = 2, alpha = 0.85) +
  coord_sf(xlim = c(bbox["xmin"], bbox["xmax"]),
           ylim = c(bbox["ymin"], bbox["ymax"]),
           expand = FALSE) +
  labs(color = "Population") +
  theme_minimal() +
   theme(axis.text.x = element_text(size = 5))

```

------------------------------------------------------------------------

## Finish and Save

```{r}
# Save to new CSV
write.csv(LM_clean, "data/LMini_iNat_Pops_filtered.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# Curating Data

------------------------------------------------------------------------

### Now that we have our data set, we need to mark iNat observations based on flower color

Lets make a copy of LMini_iNat_Pops_filtered.csv called LMini_iNat_Pops_filtered_curated.csv, and add some new columns: **polymorphic** and **color**

```{r}
## add code to do that here
```

Next, we will go through each iNat observation in our LMini_iNat_Pops_filtered_metadata.csv by clicking the hyperlink in each respective row. Then we will note the flower color in the observation's photo, choosing **pink** or **white**, and writing that down in the cell of the color column for each respective row. However, if the observation has no flowers or all flowers are still closed, we will instead write **closed** in that cell instead. Additionally, we will mark whether the observation show more than one flower color in the polymorphic column. This will be a binary variable, with **polymorphic observations receiving a 1** and **monomorphic observations receiving a 0.**

Once the **polymorphic** and **color** columns are filled out we can move on to the next step. Make sure to save it as a CSV after editing.\

------------------------------------------------------------------------

## Filtering out closed flowers

```{r}
#Read in curated CSV 
LMC <- read.csv("data/LMini_iNat_Pops_filtered_curated.csv")

LMCF <- LMC %>%
  filter(!(color1 == "closed" & color2 == "closed"))
```

\
\

```{r}
LMCF_clean <- LMCF %>%
  pivot_longer(
    cols = starts_with("color"),
    values_to = "color",
    names_to = "slot"
  ) %>%
  filter(
    !is.na(color),
    color != "",
    color != "closed"
  )
```

```{r}
pop_colors <- LMCF_clean %>%
  group_by(population_id, latitude, longitude, color) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(population_id) %>%
  mutate(prop = n / sum(n))
```

```{r}
pop_pies <- pop_colors %>%
  select(population_id, latitude, longitude, color, n) %>%
  pivot_wider(
    names_from = color,
    values_from = n,
    values_fill = 0
  )
```

\

```{r}
pop_pies <- LMCF_clean %>%
  group_by(population_id) %>%
  summarise(
    latitude  = mean(latitude),
    longitude = mean(longitude),
    white = sum(color == "white"),
    pink  = sum(color == "pink"),
    total_n = n(),
    .groups = "drop"
  )
```

\

```{r}
# World basemap
world <- ne_countries(scale = "medium", returnclass = "sf")

# Bounding box from population-level data
bbox <- c(
  xmin = min(pop_pies$longitude) - 0.5,
  xmax = max(pop_pies$longitude) + 0.5,
  ymin = min(pop_pies$latitude) - 0.5,
  ymax = max(pop_pies$latitude) + 0.5
)

# Plot
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "gray80") +
  geom_scatterpie(
    data = pop_pies,
    aes(x = longitude, y = latitude),
    cols = c("white", "pink"),
    pie_scale = 0.6,
    color = "black"
  ) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  scale_fill_manual(
    values = c(white = "white", pink = "#d81b60"),
    name = "Flower color"
  ) +
  theme_minimal()
```

```{r}
library(ggplot2)
library(scatterpie)

land <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

ggplot() +
  geom_sf(
    data = land,
    fill = "grey95",
    color = "grey70",
    linewidth = 0.2
  ) +
  geom_scatterpie(
    data = pop_pies,
    aes(x = longitude, y = latitude),
    cols = c("white", "pink"),
    pie_scale = 0.6,
    color = "black"
  ) +
  coord_sf(
    xlim = range(pop_pies$longitude) + c(-0.2, 0.2),
    ylim = range(pop_pies$latitude) + c(-0.2, 0.2),
    expand = FALSE
  ) +
  scale_x_continuous(n.breaks = 4) +
  scale_y_continuous(n.breaks = 4) +
  scale_fill_manual(
    values = c(white = "white", pink = "#d81b60"),
    name = "Flower color"
  ) +
  theme_minimal()




```

```{r}
#install.packages("prettymapr")
library(prettymapr)
library(ggspatial)

ggplot() +
  annotation_map_tile(type = "esri") +
  geom_scatterpie(
    data = pop_pies,
    aes(x = longitude, y = latitude),
    cols = c("white", "pink"),
    pie_scale = 0.6,
    color = "black"
  ) +
  coord_sf(crs = 3857) +
  scale_fill_manual(
    values = c(white = "white", pink = "#d81b60"),
    name = "Flower color"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

\
